---
title: "Exploratory Analysis on Traits"
author: "Xianwen He"
date: "2024-11-19"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
```


## Load Traits in Interest

```{r}
# load full data set
alldata <- readRDS("../Data/alldata.rds")
```


```{r}
traits.in.interest <- c('id', "PMAT24_A_CR", "ReadEng_AgeAdj", "PicVocab_AgeAdj",
                        "IWRD_TOT", "ProcSpeed_AgeAdj", "VSPLOT_TC", "SCPT_TPRT",
                        "ListSort_AgeAdj", "PicSeq_AgeAdj",
                        "SSAGA_ChildhoodConduct", "SSAGA_Alc_12_Frq_Drk", "SSAGA_TB_Reg_CPD",
                        'SSAGA_Times_Used_Illicits', 'SSAGA_Times_Used_Stimulants',
                        'Height', 'BMI', 'SSAGA_Income', 'SSAGA_Educ', 'age', 'gender',
                        'ASR_Attn_Raw', 'ASR_Totp_Raw', 'DSM_Adh_Raw')
```



## Data Screening and Processing

```{r}
demographic.data <- alldata %>% select(all_of(traits.in.interest))
```

### Cognitive Scores

```{r}
summary(demographic.data %>% select("PMAT24_A_CR", "ReadEng_AgeAdj", "PicVocab_AgeAdj",
                        "IWRD_TOT", "ProcSpeed_AgeAdj", "VSPLOT_TC", "SCPT_TPRT",
                        "ListSort_AgeAdj", "PicSeq_AgeAdj"))
```
### Psychiatric History

```{r}
table(demographic.data$SSAGA_ChildhoodConduct)

# set to binary variable
demographic.data$SSAGA_ChildhoodConduct_binary <- factor(demographic.data$SSAGA_ChildhoodConduct > 0, levels=c(F, T))
table(demographic.data$SSAGA_ChildhoodConduct_binary)
```


### Substance Abuse

`SSAGA_Alc_12_Frq_Drk` and `SSAGA_TB_Reg_CPD`
```{r}
table(demographic.data$SSAGA_Alc_12_Frq_Drk)
table(demographic.data$SSAGA_TB_Reg_CPD)
sum(is.na(demographic.data$SSAGA_TB_Reg_CPD))  # too many NAs. Discard this variable
```
```{r}
# the frequency of being drunk
drk_ordered <-  rep(NA, nrow(demographic.data))
drk_ordered[demographic.data$SSAGA_Alc_12_Frq_Drk==1] <- 1  # male-low-freq
drk_ordered[demographic.data$SSAGA_Alc_12_Frq_Drk==2] <- 1  # female-low-freq
drk_ordered[demographic.data$SSAGA_Alc_12_Frq_Drk==3] <- 2  # high-freq
drk_ordered[demographic.data$SSAGA_Alc_12_Frq_Drk==4] <- 0  # never
demographic.data$SSAGA_Alc_12_Frq_Drk_ordered <- factor(drk_ordered, levels=c(0, 1, 2), ordered=TRUE)

# whether or not drunk
demographic.data$SSAGA_Alc_12_Frq_Drk_binary <- factor(demographic.data$SSAGA_Alc_12_Frq_Drk<4, levels=c(F, T))
```


`SSAGA_Times_Used_Illicits` and `SSAGA_Times_Used_Stimulants`
```{r}
table(demographic.data$SSAGA_Times_Used_Illicits)
table(demographic.data$SSAGA_Times_Used_Stimulants)
```

```{r}
# convert to binary for data balance
demographic.data$SSAGA_Times_Used_Illicits_binary <- factor(demographic.data$SSAGA_Times_Used_Illicits > 0, levels=c(F, T))
demographic.data$SSAGA_Times_Used_Stimulants_binary <- factor(demographic.data$SSAGA_Times_Used_Stimulants > 0, levels=c(F, T))

summary(demographic.data %>% select(SSAGA_Times_Used_Illicits_binary, SSAGA_Times_Used_Stimulants_binary))
```

### Attention Scores

`ASR_Attn_Raw`, `ASR_Totp_Raw`, and `DSM_Adh_Raw`.
```{r}
summary(demographic.data %>% select('ASR_Attn_Raw', 'ASR_Totp_Raw', 'DSM_Adh_Raw'))
```

### Demographic and Health Information

`Height` and `BMI`
```{r}
summary(demographic.data %>% select('Height', 'BMI'))
```
`SSAGA_Income` and `SSAGA_Educ`

```{r}
table(demographic.data$SSAGA_Income)
table(demographic.data$SSAGA_Educ)
```

Convert `SSAGA_Income` and `SSAGA_Educ` into ordered categorical variable.
```{r}
demographic.data$SSAGA_Income_ordered <- factor(demographic.data$SSAGA_Income, 
                                                levels = 1:8, ordered=T)

demographic.data$SSAGA_Educ_ordered <- factor(demographic.data$SSAGA_Educ,
                                      levels=11:17, ordered=T)
```


`Age` and `Gender`
```{r}
summary(demographic.data$age)
```

```{r}
table(demographic.data$gender)
```
```{r}
demographic.data$gender_factor <- factor(demographic.data$gender-1, levels=c(0, 1), ordered=FALSE)
table(demographic.data$gender_factor)
```


## Exploratory Analysis

### Attention Scores

```{r}
ggplot(demographic.data, aes(x = ASR_Attn_Raw, y = DSM_Adh_Raw)) +
  geom_point(size = 3) +
  labs(
    title = "ASR_Attn_Raw v.s. DSM_Adh_Raw",
    x = "ASR_Attn_Raw", y = "DSM_Adh_Raw") +
  theme_minimal(base_size=12)
```

```{r}
ggplot(demographic.data, aes(x = ASR_Totp_Raw, y = DSM_Adh_Raw)) +
  geom_point(size = 3) +
  labs(
    title = "ASR_Totp_Raw v.s. DSM_Adh_Raw",
    x = "ASR_Totp_Raw", y = "DSM_Adh_Raw") +
  theme_minimal(base_size=12)
```

In the following texts, we are going to focus on `DSM_Adh_Raw`.

```{r}
ggplot(demographic.data, aes(x = DSM_Adh_Raw)) +
  geom_histogram(binwidth = 2, fill = "lightblue", color = "blue", alpha=0.7) +
  labs(title = "Histogram of DSM_Adh_Raw", x = "", y = "") +
  theme_minimal(base_size=12)

ggplot(demographic.data, aes(x = log(DSM_Adh_Raw+1))) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "blue", alpha=0.7) +
  labs(title = "Histogram of DSM_Adh_Raw", x = "", y = "") +
  theme_minimal(base_size=12)
```
```{r}
demographic.data$DSM_Adh_Raw_log <- log(demographic.data$DSM_Adh_Raw+1)
```


### Substance Abuse

**SSAGA_Times_Used_Illicits_binary**

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Times_Used_Illicits_binary)), aes(x = SSAGA_Times_Used_Illicits_binary, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_Times_Used_Illicits",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

**SSAGA_Times_Used_Stimulants_binary**


```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Times_Used_Stimulants_binary)), aes(x = SSAGA_Times_Used_Stimulants_binary, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_Times_Used_Stimulants",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

**SSAGA_Alc_12_Frq_Drk_ordered**

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Alc_12_Frq_Drk_ordered)), aes(x = SSAGA_Alc_12_Frq_Drk_ordered, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_Alc_12_Frq_Drk_ordered",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Alc_12_Frq_Drk_binary)), aes(x = SSAGA_Alc_12_Frq_Drk_binary, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "ADSM_Adh_Raw by SSAGA_Alc_12_Frq_Drk_binary",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

Maintain `SSAGA_Alc_12_Frq_Drk_binary`.


### Phychiatric History

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_ChildhoodConduct_binary)), aes(x = SSAGA_ChildhoodConduct_binary, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_ChildhoodConduct_binary",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

### Demographic Information

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Educ_ordered)), aes(x = SSAGA_Educ_ordered, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_Educ_ordered",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

```{r}
ggplot(demographic.data %>% filter(!is.na(SSAGA_Income_ordered)), aes(x = SSAGA_Income_ordered, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by SSAGA_Income_ordered",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

Education and family income cannot distinguish the DSM scores.


**BMI** and **height**

```{r}
ggplot(demographic.data, aes(x = BMI, y = DSM_Adh_Raw)) +
  geom_point(size = 3) +
  labs(
    title = "BMI v.s. DSM_Adh_Raw",
    x = "BMI", y = "DSM_Adh_Raw") +
  theme_minimal(base_size=12)
```

```{r}
ggplot(demographic.data, aes(x = Height, y = DSM_Adh_Raw)) +
  geom_point(size = 3) +
  labs(
    title = "Height v.s. DSM_Adh_Raw",
    x = "Height", y = "DSM_Adh_Raw") +
  theme_minimal(base_size=12)
```

**gender**

```{r}
ggplot(demographic.data, aes(x = gender_factor, y = DSM_Adh_Raw)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs( 
    title = "DSM_Adh_Raw by gender",
    x = "", y = "") + 
  theme_minimal(base_size=12)
```

**Age**
```{r}
ggplot(demographic.data, aes(x = age, y = DSM_Adh_Raw)) +
  geom_point(size = 3) +
  labs(
    title = "age v.s. DSM_Adh_Raw",
    x = "age", y = "DSM_Adh_Raw") +
  theme_minimal(base_size=12)
```

### Cognitive Scores

```{r}
cogn.dat <- demographic.data %>% select("PMAT24_A_CR", "ReadEng_AgeAdj", "PicVocab_AgeAdj",
                        "IWRD_TOT", "ProcSpeed_AgeAdj", "VSPLOT_TC", "SCPT_TPRT",
                        "ListSort_AgeAdj", "PicSeq_AgeAdj", 'DSM_Adh_Raw', 'DSM_Adh_Raw_log')
cogn.dat <- na.omit(cogn.dat)
```

```{r}
cogn.lm <- lm(DSM_Adh_Raw~PMAT24_A_CR+ReadEng_AgeAdj+PicVocab_AgeAdj+IWRD_TOT+ProcSpeed_AgeAdj+VSPLOT_TC+SCPT_TPRT+ListSort_AgeAdj+PicSeq_AgeAdj, data=cogn.dat)
cogn.log.lm <- lm(DSM_Adh_Raw_log~PMAT24_A_CR+ReadEng_AgeAdj+PicVocab_AgeAdj+IWRD_TOT+ProcSpeed_AgeAdj+VSPLOT_TC+SCPT_TPRT+ListSort_AgeAdj+PicSeq_AgeAdj, data=cogn.dat)
```


```{r}
summary(cogn.lm)
```

```{r}
summary(cogn.log.lm )
```

### Save Data

Save the converted demographic data set.
```{r}
save(demographic.data, file='../Data/demographic.data.RData')
```


## Modeling

### Data for Modeling

```{r}
traits.log.dat <- demographic.data %>% select('id', "PMAT24_A_CR", "ReadEng_AgeAdj", "PicVocab_AgeAdj",
                        "IWRD_TOT", "ProcSpeed_AgeAdj", "VSPLOT_TC", "SCPT_TPRT",
                        "ListSort_AgeAdj", "PicSeq_AgeAdj",
                        "SSAGA_ChildhoodConduct_binary", "SSAGA_Alc_12_Frq_Drk_binary",
                        'SSAGA_Times_Used_Illicits_binary', 'SSAGA_Times_Used_Stimulants_binary',
                        'BMI', 'gender_factor',
                        'DSM_Adh_Raw_log')

traits.dat <- demographic.data %>% select('id', "PMAT24_A_CR", "ReadEng_AgeAdj", "PicVocab_AgeAdj",
                        "IWRD_TOT", "ProcSpeed_AgeAdj", "VSPLOT_TC", "SCPT_TPRT",
                        "ListSort_AgeAdj", "PicSeq_AgeAdj",
                        "SSAGA_ChildhoodConduct_binary", "SSAGA_Alc_12_Frq_Drk_binary",
                        'SSAGA_Times_Used_Illicits_binary', 'SSAGA_Times_Used_Stimulants_binary',
                        'BMI', 'gender_factor',
                        'DSM_Adh_Raw')

traits.dat <- na.omit(traits.dat)
traits.log.dat <- na.omit(traits.log.dat)
```

```{r}
save(traits.log.dat, file='../Data/traits.log.dat.RData')
save(traits.dat, file='../Data/traits.dat.RData')
```


### Linear Regression

```{r}
traits.log.lm <- lm(DSM_Adh_Raw_log~., data=traits.log.dat[, -1])
traits.lm <- lm(DSM_Adh_Raw~., data=traits.dat[, -1])
```

```{r}
summary(traits.log.lm)
```

```{r}
summary(traits.lm)
```



